---
title: "Hazard plots with left censored data"
output: html_notebook
---

This file plots the baseline hazards for each of the 7 transitions of interest against the 
intercept-only parametric distributions and compares AIC values for the fits. 
For each transition, Kaplan-Meier curves are also plotted against the parametric intercept-only
curves of the best-fitting distribution(s) for visual inspection. 

based on bshazard: A Flexible Tool for Nonparametric Smoothing of the Hazard Function
https://journal.r-project.org/archive/2014/RJ-2014-028/RJ-2014-028.pdf 

parametric distributions from: # from https://devinincerti.com/2019/06/18/parametric_survival.html


```{r}
library(data.table)
data.table::setDTthreads(10) #this is so that don't use all the processors
library(fst)
fst::threads_fst(10)
library(qs)
library(flexsurv)
library(ggplot2)
library(future.apply)
library(future)
library(splines)
library(ggthemes)
options(future.fork.enable = TRUE) # TODO remove for production
options(future.rng.onMisuse = "ignore") # Remove false warning
plan(multicore(workers = 10))

data_dir_PM <-
  function(x = character(0))
    paste0("/mnt/", Sys.info()[["user"]],
           "/UoL/CPRD2019mm/forthesis/", x)

data_dir_Diseaselist <-
  function(x = character(0))
    paste0("/mnt/", Sys.info()[["user"]],
           "/UoL/CPRD2019mm/Disease_lists/", x)


data_dir_lookup <-
  function(x = character(0))
    paste0("/mnt/", Sys.info()[["user"]],
           "/UoL/CPRD2019mm/Dictionaries etc/", x)

output_path_plot <- 
     function(x = character(0))
    paste0("/mnt/", Sys.info()[["user"]],
           "/UoL/CPRD2019mm/forthesis/bshazplot/", x)

myDTmstate_age <- read_fst(data_dir_PM("accumDTmstate_age.fst"), as.data.table = TRUE)[
  yob >= 1919 & yob <= 1986 & imd!= "" & region != "" & gender != "I"]





dists <- c("exp", "weibull", "gompertz", "gamma",
           "lognormal", "llogis", "gengamma")
dists_long <- c("Exponential", "Weibull (AFT)",
                "Gompertz", "Gamma", "Lognormal", "Log-logistic",
                "Generalized gamma")
n_dists <- length(dists)

foo <- function(dist, data) {
flexsurvreg(Surv(Tstart2, Tstop2, status) ~ 1,
              data = data,
              dist = dist)
}
fooa <- function(dist, data) {
  flexsurvreg(Surv(Tstart2, Tstop2, status) ~ log(yob),
              data = data,
              dist = dist)
}

time10 <- c(0,10,20,30,40,50,60,70,80)
theme_set(new = theme_few())
theme_update(legend.position="bottom")
```


# T1 healthy -> 1 cond
```{r, warning=FALSE, message=FALSE}
library(bshazard)
j <- 1L

models <- future_lapply(dists, foo, myDTmstate_age[trans == j])
AICs <- sapply(models, `[[`, "AIC")
names(AICs) <- dists
sort(AICs)
names(sort(AICs)[1])
args(paste0("d", names(sort(AICs)[1]))) #gengamma
parametric_haz <- vector(mode = "list", length = length(dists))
for (i in 1:length(dists)){
  parametric_haz[[i]] <- summary(models[[i]], type = "hazard",
                                 ci = FALSE, tidy = TRUE)
  parametric_haz[[i]]$method <- dists_long[i]
}
parametric_haz <- rbindlist(parametric_haz)


bs_haz_est <- bshazard(Surv(Tstart2, Tstop2, status) ~ 1, 
                              data = myDTmstate_age[trans == j])

bs_haz <- data.table(time = bs_haz_est$time,
                         est = bs_haz_est$hazard,
                         method = "B-spline hazard")
haz <- rbind(bs_haz, parametric_haz)
haz[, method := factor(method,
                       levels = c("B-spline hazard",
                                  dists_long))]
n_dists <- length(dists)
ggplot(haz, aes(x = time, y = est, col = method, linetype = method)) +
  geom_line() +
  xlab("Years") + ylab("Hazard") +
  scale_colour_manual(name = "",
                      values = c("black", rainbow(n_dists))) +
  scale_linetype_manual(name = "",
                        values = c(1, rep_len(2:6, n_dists))) 
  ggsave(filename = output_path_plot(paste0("bshaz_lc_t", j,".png",sep="")))


kmfit <- survfit(Surv(Tstart2, Tstop2, status) ~ 1, 
                 data = myDTmstate_age[trans == j])

summary(kmfit, times = time10)
myDTmstate_age[trans == j, .(totalatrisk = .N, totalevents = sum(status ==1) )]

```
## Fitted survival curves
```{r}
plot(models[[3]], xlab = "Age in years", ylab = "Overall survival probability")
lines(models[[7]], col = "blue")
legend("bottomleft",c("gompertz", "gengamma"),
       col = c("red", "blue"), lty = 1)
title("Transition from healthy -> 1 condition, \n KM curve & intercept only parametric models")

```

# T2 healthy -> Death
```{r, warning=FALSE}
j <- 2L

models <- future_lapply(dists, foo, myDTmstate_age[trans == j])
AICs <- sapply(models, `[[`, "AIC")
names(AICs) <- dists
sort(AICs)
names(sort(AICs)[1])
args(paste0("d", names(sort(AICs)[1]))) #THIS HAS CHANGED TO GENGAMMA...
#BUT, the Gompertz fits better in the plots later on, so Gompertz probably makes more sense

parametric_haz <- vector(mode = "list", length = length(dists))
for (i in 1:length(dists)){
  parametric_haz[[i]] <- summary(models[[i]], type = "hazard",
                                 ci = FALSE, tidy = TRUE)
  parametric_haz[[i]]$method <- dists_long[i]
}
parametric_haz <- rbindlist(parametric_haz)

bs_haz_est <- bshazard(Surv(Tstart2, Tstop2, status) ~ 1, 
                              data = myDTmstate_age[trans == j])

bs_haz <- data.table(time = bs_haz_est$time,
                         est = bs_haz_est$hazard,
                         method = "B-spline hazard")
haz <- rbind(bs_haz, parametric_haz)
haz[, method := factor(method,
                       levels = c("B-spline hazard",
                                  dists_long))]
n_dists <- length(dists)
ggplot(haz, aes(x = time, y = est, col = method, linetype = method)) +
  geom_line() +
  xlab("Years") + ylab("Hazard") +
  scale_colour_manual(name = "",
                      values = c("black", rainbow(n_dists))) +
  scale_linetype_manual(name = "",
                        values = c(1, rep_len(2:6, n_dists))) 
ggsave(filename = output_path_plot(paste0("bshaz_lc_t", j,".png",sep="")))


kmfit <- survfit(Surv(Tstart2, Tstop2, status) ~ 1, 
                 data = myDTmstate_age[trans == j])

summary(kmfit, times = time10)
myDTmstate_age[trans == j, .(totalatrisk = .N, totalevents = sum(status ==1) )]


```

## Fitted survival curves
```{r}
plot(models[[3]], xlab = "Age in years", ylab = "Overall survival probability")
lines(models[[7]], col = "blue")
legend("bottomleft",c("gompertz", "gengamma"),
       col = c("red", "blue"), lty = 1)
title("Transition from healthy -> death, \n KM curve & intercept only parametric models")

```

# T3 1 Cond -> BMM
```{r, warning=FALSE}
j <- 3L

models <- future_lapply(dists, foo, myDTmstate_age[trans == j])
AICs <- sapply(models, `[[`, "AIC")
names(AICs) <- dists
print(sort(AICs))
names(sort(AICs)[1])
args(paste0("d", names(sort(AICs)[1]))) #gengamma

parametric_haz <- vector(mode = "list", length = length(dists))
for (i in 1:length(dists)){
  parametric_haz[[i]] <- summary(models[[i]], type = "hazard",
                                 ci = FALSE, tidy = TRUE)
  parametric_haz[[i]]$method <- dists_long[i]
}
parametric_haz <- rbindlist(parametric_haz)

bs_haz_est <- bshazard(Surv(Tstart2, Tstop2, status) ~ 1, 
                              data = myDTmstate_age[trans == j])

bs_haz <- data.table(time = bs_haz_est$time,
                         est = bs_haz_est$hazard,
                         method = "B-spline hazard")
haz <- rbind(bs_haz, parametric_haz)
haz[, method := factor(method,
                       levels = c("B-spline hazard",
                                  dists_long))]
n_dists <- length(dists)
ggplot(haz, aes(x = time, y = est, col = method, linetype = method)) +
  geom_line() +
  xlab("Years") + ylab("Hazard") +
  scale_colour_manual(name = "",
                      values = c("black", rainbow(n_dists))) +
  scale_linetype_manual(name = "",
                        values = c(1, rep_len(2:6, n_dists))) 
  ggsave(filename = output_path_plot(paste0("bshaz_lc_t", j,".png",sep="")))


kmfit <- survfit(Surv(Tstart2, Tstop2, status) ~ 1, 
                 data = myDTmstate_age[trans == j])

summary(kmfit, times = time10)
myDTmstate_age[trans == j, .(totalatrisk = .N, totalevents = sum(status ==1) )]


```


## Fitted survival curves
```{r}
plot(models[[3]], xlab = "Age in years", ylab = "Overall survival probability")
lines(models[[7]], col = "blue")
legend("bottomleft",c("gompertz", "gengamma"),
       col = c("red", "blue"), lty = 1)
title("Transition from 1 cond -> bmm, \n KM curve & intercept only parametric models")

```

# T4 1 Cond -> Death
```{r, warning=FALSE }
j <- 4L

models <- future_lapply(dists, foo, myDTmstate_age[trans == j])
AICs <- sapply(models, `[[`, "AIC")
names(AICs) <- dists
print(sort(AICs))
names(sort(AICs)[1])
args(paste0("d", names(sort(AICs)[1]))) #gengamma, but maybe affected by small numbers on the right
#gengamma  gompertz   weibull    llogis     gamma lognormal       exp 
# 26177.03  35242.78  36205.01  36478.27  37016.19  37867.57  40690.19 

parametric_haz <- vector(mode = "list", length = length(dists))
for (i in 1:length(dists)){
  parametric_haz[[i]] <- summary(models[[i]], type = "hazard",
                                 ci = FALSE, tidy = TRUE)
  parametric_haz[[i]]$method <- dists_long[i]
}
parametric_haz <- rbindlist(parametric_haz)

bs_haz_est <- bshazard(Surv(Tstart2, Tstop2, status) ~ 1, 
                              data = myDTmstate_age[trans == j])

bs_haz <- data.table(time = bs_haz_est$time,
                         est = bs_haz_est$hazard,
                         method = "B-spline hazard")
haz <- rbind(bs_haz, parametric_haz)
haz[, method := factor(method,
                       levels = c("B-spline hazard",
                                  dists_long))]
n_dists <- length(dists)
ggplot(haz, aes(x = time, y = est, col = method, linetype = method)) +
  geom_line() +
  xlab("Years") + ylab("Hazard") +
  scale_colour_manual(name = "",
                      values = c("black", rainbow(n_dists))) +
  scale_linetype_manual(name = "",
                        values = c(1, rep_len(2:6, n_dists))) 
 ggsave(filename = output_path_plot(paste0("bshaz_lc_t", j,".png",sep="")))


kmfit <- survfit(Surv(Tstart2, Tstop2, status) ~ 1, 
                 data = myDTmstate_age[trans == j])

summary(kmfit, times = time10)
myDTmstate_age[trans == j, .(totalatrisk = .N, totalevents = sum(status ==1) )]


```

## Fitted survival curves
```{r}
plot(models[[3]], xlab = "Age in years", ylab = "Overall survival probability")
lines(models[[7]], col = "blue")
legend("bottomleft",c("gompertz", "gengamma"),
       col = c("red", "blue"), lty = 1)
title("Transition from 1 cond -> death, \n KM curve & intercept only parametric models")

```


# T5 BMM -> CMM
```{r, warning=FALSE }
j <- 5L

models <- future_lapply(dists, foo, myDTmstate_age[trans == j])
AICs <- sapply(models, `[[`, "AIC")
names(AICs) <- dists
sort(AICs)
names(sort(AICs)[1])
args(paste0("d", names(sort(AICs)[1]))) #gengamma

parametric_haz <- vector(mode = "list", length = length(dists))
for (i in 1:length(dists)){
  parametric_haz[[i]] <- summary(models[[i]], type = "hazard",
                                 ci = FALSE, tidy = TRUE)
  parametric_haz[[i]]$method <- dists_long[i]
}
parametric_haz <- rbindlist(parametric_haz)

bs_haz_est <- bshazard(Surv(Tstart2, Tstop2, status) ~ 1, 
                              data = myDTmstate_age[trans == j])

bs_haz <- data.table(time = bs_haz_est$time,
                         est = bs_haz_est$hazard,
                         method = "B-spline hazard")
haz <- rbind(bs_haz, parametric_haz)
haz[, method := factor(method,
                       levels = c("B-spline hazard",
                                  dists_long))]
n_dists <- length(dists)
ggplot(haz, aes(x = time, y = est, col = method, linetype = method)) +
  geom_line() +
  xlab("Years") + ylab("Hazard") +
  scale_colour_manual(name = "",
                      values = c("black", rainbow(n_dists))) +
  scale_linetype_manual(name = "",
                        values = c(1, rep_len(2:6, n_dists))) 
  ggsave(filename = output_path_plot(paste0("bshaz_lc_t", j,".png",sep="")))


kmfit <- survfit(Surv(Tstart2, Tstop2, status) ~ 1, 
                 data = myDTmstate_age[trans == j])

summary(kmfit, times = time10)
myDTmstate_age[trans == j, .(totalatrisk = .N, totalevents = sum(status ==1) )]


```

## Fitted survival curves
```{r}
plot(models[[3]], xlab = "Age in years", ylab = "Overall survival probability")
lines(models[[7]], col = "blue")
legend("bottomleft",c("gompertz", "gengamma"),
       col = c("red", "blue"), lty = 1)
title("Transition from bmm -> cmm, \n KM curve & intercept only parametric models")

# the gompertz does look better though..
```


# T6 BMM -> Death
```{r, warning=FALSE }
j <- 6L

models <- future_lapply(dists, foo, myDTmstate_age[trans == j])
AICs <- sapply(models, `[[`, "AIC")
names(AICs) <- dists
sort(AICs)
names(sort(AICs)[1])
args(paste0("d", names(sort(AICs)[1]))) #gompertz

parametric_haz <- vector(mode = "list", length = length(dists))
for (i in 1:length(dists)){
  parametric_haz[[i]] <- summary(models[[i]], type = "hazard",
                                 ci = FALSE, tidy = TRUE)
  parametric_haz[[i]]$method <- dists_long[i]
}
parametric_haz <- rbindlist(parametric_haz)

bs_haz_est <- bshazard(Surv(Tstart2, Tstop2, status) ~ 1, 
                              data = myDTmstate_age[trans == j])

bs_haz <- data.table(time = bs_haz_est$time,
                         est = bs_haz_est$hazard,
                         method = "B-spline hazard")
haz <- rbind(bs_haz, parametric_haz)
haz[, method := factor(method,
                       levels = c("B-spline hazard",
                                  dists_long))]
n_dists <- length(dists)
ggplot(haz, aes(x = time, y = est, col = method, linetype = method)) +
  geom_line() +
  xlab("Years") + ylab("Hazard") +
  scale_colour_manual(name = "",
                      values = c("black", rainbow(n_dists))) +
  scale_linetype_manual(name = "",
                        values = c(1, rep_len(2:6, n_dists))) 
  ggsave(filename = output_path_plot(paste0("bshaz_lc_t", j,".png",sep="")))


kmfit <- survfit(Surv(Tstart2, Tstop2, status) ~ 1, 
                 data = myDTmstate_age[trans == j])

summary(kmfit, times = time10)
myDTmstate_age[trans == j, .(totalatrisk = .N, totalevents = sum(status ==1) )]

```

## Fitted survival curves
```{r}
plot(models[[3]], xlab = "Age in years", ylab = "Overall survival probability")
lines(models[[7]], col = "blue")
legend("bottomleft",c("gompertz", "gengamma"),
       col = c("red", "blue"), lty = 1)
title("Transition from bmm -> death, \n KM curve & intercept only parametric models")

```



# T7 CMM -> Death
```{r, warning=FALSE }
j <- 7L
AICs <- sapply(models, `[[`, "AIC")
names(AICs) <- dists
sort(AICs)
names(sort(AICs)[1])
args(paste0("d", names(sort(AICs)[1])))

models <- future_lapply(dists, foo, myDTmstate_age[trans == j]) #gompertz

parametric_haz <- vector(mode = "list", length = length(dists))
for (i in 1:length(dists)){
  parametric_haz[[i]] <- summary(models[[i]], type = "hazard",
                                 ci = FALSE, tidy = TRUE)
  parametric_haz[[i]]$method <- dists_long[i]
}
parametric_haz <- rbindlist(parametric_haz)

bs_haz_est <- bshazard(Surv(Tstart2, Tstop2, status) ~ 1, 
                              data = myDTmstate_age[trans == j])

bs_haz <- data.table(time = bs_haz_est$time,
                         est = bs_haz_est$hazard,
                         method = "B-spline hazard")
haz <- rbind(bs_haz, parametric_haz)
haz[, method := factor(method,
                       levels = c("B-spline hazard",
                                  dists_long))]
n_dists <- length(dists)
ggplot(haz, aes(x = time, y = est, col = method, linetype = method)) +
  geom_line() +
  xlab("Years") + ylab("Hazard") +
  scale_colour_manual(name = "",
                      values = c("black", rainbow(n_dists))) +
  scale_linetype_manual(name = "",
                        values = c(1, rep_len(2:6, n_dists))) 
  ggsave(filename = output_path_plot(paste0("bshaz_lc_t", j,".png",sep="")))


kmfit <- survfit(Surv(Tstart2, Tstop2, status) ~ 1, 
                 data = myDTmstate_age[trans == j])

summary(kmfit, times = time10)
myDTmstate_age[trans == j, .(totalatrisk = .N, totalevents = sum(status ==1) )]


```

## Fitted survival curves
```{r}
plot(models[[3]], xlab = "Age in years", ylab = "Overall survival probability")
lines(models[[7]], col = "blue")
legend("bottomleft",c("gompertz", "gengamma"),
       col = c("red", "blue"), lty = 1)
title("Transition from cmm -> death, \n KM curve & intercept only parametric models")

```


# T1 healthy -> 1 cond -h18
```{r, warning=FALSE, message=FALSE}
library(bshazard)
j <- 1L
myDTmstate_age[from ==1, Tstart2 := 0]
models <- future_lapply(dists, foo, myDTmstate_age[trans == j])
AICs <- sapply(models, `[[`, "AIC")
names(AICs) <- dists
sort(AICs)
names(sort(AICs)[1])
args(paste0("d", names(sort(AICs)[1]))) #gengamma

parametric_haz <- vector(mode = "list", length = length(dists))
for (i in 1:length(dists)){
  parametric_haz[[i]] <- summary(models[[i]], type = "hazard",
                                 ci = FALSE, tidy = TRUE)
  parametric_haz[[i]]$method <- dists_long[i]
}
parametric_haz <- rbindlist(parametric_haz)


bs_haz_est <- bshazard(Surv(Tstart2, Tstop2, status) ~ 1, 
                              data = myDTmstate_age[trans == j])

bs_haz <- data.table(time = bs_haz_est$time,
                         est = bs_haz_est$hazard,
                         method = "B-spline hazard")
haz <- rbind(bs_haz, parametric_haz)
haz[, method := factor(method,
                       levels = c("B-spline hazard",
                                  dists_long))]
n_dists <- length(dists)
ggplot(haz, aes(x = time, y = est, col = method, linetype = method)) +
  geom_line() +
  xlab("Years") + ylab("Hazard") +
  scale_colour_manual(name = "",
                      values = c("black", rainbow(n_dists))) +
  scale_linetype_manual(name = "",
                        values = c(1, rep_len(2:6, n_dists))) 
  ggsave(filename = output_path_plot(paste0("bshaz_lc_h_18_t", j,".png",sep="")))


kmfit <- survfit(Surv(Tstart2, Tstop2, status) ~ 1, 
                 data = myDTmstate_age[trans == j])

summary(kmfit, times = time10)
myDTmstate_age[trans == j, .(totalatrisk = .N, totalevents = sum(status ==1) )]

```

## Fitted survival curves
```{r}
plot(models[[3]], xlab = "Age in years", ylab = "Overall survival probability")
lines(models[[7]], col = "blue")
legend("bottomleft",c("gompertz", "gengamma"),
       col = c("red", "blue"), lty = 1)
title("Transition from healthy -> 1condition (start at 18), \n KM curve & intercept only parametric models")

```

# T2 healthy -> Death - h18
```{r, warning=FALSE}
j <- 2L

models <- future_lapply(dists, foo, myDTmstate_age[trans == j])
AICs <- sapply(models, `[[`, "AIC")
names(AICs) <- dists
sort(AICs)
names(sort(AICs)[1])
args(paste0("d", names(sort(AICs)[1]))) #gompertz

parametric_haz <- vector(mode = "list", length = length(dists))
for (i in 1:length(dists)){
  parametric_haz[[i]] <- summary(models[[i]], type = "hazard",
                                 ci = FALSE, tidy = TRUE)
  parametric_haz[[i]]$method <- dists_long[i]
}
parametric_haz <- rbindlist(parametric_haz)

bs_haz_est <- bshazard(Surv(Tstart2, Tstop2, status) ~ 1, 
                              data = myDTmstate_age[trans == j])

bs_haz <- data.table(time = bs_haz_est$time,
                         est = bs_haz_est$hazard,
                         method = "B-spline hazard")
haz <- rbind(bs_haz, parametric_haz)
haz[, method := factor(method,
                       levels = c("B-spline hazard",
                                  dists_long))]
n_dists <- length(dists)
ggplot(haz, aes(x = time, y = est, col = method, linetype = method)) +
  geom_line() +
  xlab("Years") + ylab("Hazard") +
  scale_colour_manual(name = "",
                      values = c("black", rainbow(n_dists))) +
  scale_linetype_manual(name = "",
                        values = c(1, rep_len(2:6, n_dists))) 
  ggsave(filename = output_path_plot(paste0("bshaz_lc_h_18_t", j,".png",sep="")))


kmfit <- survfit(Surv(Tstart2, Tstop2, status) ~ 1, 
                 data = myDTmstate_age[trans == j])

summary(kmfit, times = time10)
myDTmstate_age[trans == j, .(totalatrisk = .N, totalevents = sum(status ==1) )]


```

## Fitted survival curves
```{r}
plot(models[[3]], xlab = "Age in years", ylab = "Overall survival probability")
lines(models[[7]], col = "blue")
legend("bottomleft",c("gompertz", "gengamma"),
       col = c("red", "blue"), lty = 1)
title("Transition from healthy -> death (start at 18), \n KM curve & intercept only parametric models")

```
